---
title: "File-based vs database-based workflows"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## TL;DR

`duckdbfs` is designed around a **file-first** workflow: treat parquet / CSV
/ spatial files (local or remote) as the source of truth, build a lightweight
DuckDB **VIEW** over them, and query that with `dplyr`. You almost never need
to import data into a DuckDB *table* first. This is a deliberate inversion of
the classic database mental model, and it's worth explaining why — and when
to break the rule.

## Files + VIEWs (the default)

```{r}
library(duckdbfs)
library(dplyr)

tbl <- open_dataset("s3://some-bucket/data.parquet")
tbl |>
  filter(year == 2024) |>
  summarise(mean_temp = mean(temp))
```

What happens here:

- `open_dataset()` creates a DuckDB VIEW pointing at the remote parquet file.
  Nothing is downloaded; no table is materialised.
- The VIEW takes ~zero disk space and no RAM.
- `dplyr` verbs translate to SQL via `dbplyr`; the query is pushed to DuckDB,
  which streams only the bytes it needs (parquet column chunks, S3 range
  requests, HTTP byte ranges).

This pattern scales to datasets much larger than RAM and larger than local
disk, because the data never leaves its source.

## Databases + TABLES (the escape hatch)

Importing remote data into a local DuckDB table looks natural to users coming
from PostgreSQL or SQLite, but for most cloud-native data workloads it is a
pessimisation:

- It downloads the *entire* dataset (not just the columns or row groups you
  need).
- It decompresses and re-encodes into DuckDB's native format, often blowing
  up disk footprint.
- On very large datasets it can exhaust RAM or `/tmp` during the import,
  crashing the machine.

There are, however, real cases where a local table *is* the right call:

1. **You're going to query the same data many times on a slow network.** One
   one-time download amortises across many queries.
2. **You need an index DuckDB can only build on stored tables.** The spatial
   extension's R-tree index (`CREATE INDEX ... USING RTREE`) is the canonical
   example.
3. **You need a persistent on-disk database that other tools will share.**

For (1) and (2), use `mode = "TABLE"` in `open_dataset()` and point
`cached_connection()` at a disk location so DuckDB doesn't try to hold the
table in RAM:

```{r}
con <- cached_connection("/path/to/storage")
open_dataset("s3://bucket/big-data", conn = con, mode = "TABLE")
```

For (3), manage the connection yourself:

```{r}
con <- DBI::dbConnect(duckdb::duckdb(), dbdir = "shared.duckdb")
open_dataset("local/data.parquet", conn = con)
DBI::dbExecute(con, "CREATE TABLE cached AS SELECT * FROM ...")
```

## The sticky-cache gotcha

`cached_connection()` is cached *by design* — it caches the first connection
it makes and returns that same object on every subsequent call, so internal
duckdbfs functions share one connection. The trap: if the first call happened
implicitly (e.g. inside some earlier `open_dataset()` call that created the
default in-memory connection), a *later* call like
`cached_connection(dbdir = "my.duckdb")` will silently ignore the `dbdir`
argument and return the cached in-memory connection.

If you want a disk-backed connection, either:

- Call `cached_connection(dbdir = ...)` **before any other duckdbfs
  operation**, or
- Call `close_connection()` to drop the cached one, then re-create with
  your desired `dbdir`.

This was the source of confusion in
[#49](https://github.com/cboettig/duckdbfs/issues/49).

## The "R user" framing

Most R users come to data work through the
`read_csv() |> dplyr::... |> write_csv()` pipeline — data lives in *files*,
is loaded in memory for a session, then written back out. `duckdbfs` tries
to preserve that mental model but extend it to datasets that don't fit in
RAM, and to remote sources that can't be read by base-R connections. The
`fs` in `duckdbfs` stands for *filesystem*.

Tools like [ibis](https://ibis-project.org) (Python) use an identical design
on top of DuckDB: open a file, get a lazy relation, push computation down via
SQL, collect only what you need. If you already think in that idiom, the
duckdbfs defaults should feel familiar. If you're coming from a
database-administration background, think of `open_dataset()` as producing
a VIEW rather than a TABLE — and reach for TABLE mode only when you need
what a TABLE gives you (indexing, persistence, multi-process access).
